{\label{08_02_ex_24} Break the Harmonic Series into the sum of the odd and even terms:
$$\sum_{n=1}^\infty \frac1n = \sum_{n=1}^\infty \frac{1}{2n-1}+\sum_{n=1}^\infty \frac{1}{2n}.$$ The goal is to show that each of the series on the right diverge.
		\begin{enumerate}
		\item		Show why $\ds \sum_{n=1}^\infty \frac{1}{2n-1}>\sum_{n=1}^\infty \frac{1}{2n}$. 
		
		(Compare each $n^\text{th}$ partial sum.)
		\item		Show why $\ds\sum_{n=1}^\infty \frac{1}{2n-1}<1+\sum_{n=1}^\infty \frac{1}{2n}$
		\item		Explain why (a) and (b) demonstrate that the series of odd terms is convergent, if, and only if, the series of even terms is also convergent. (That is, show both converge or both diverge.)
		\item		Explain why knowing the Harmonic Series is divergent determines that the even and odd series are also divergent.
		\end{enumerate}
}
{\begin{enumerate}
\item		The $n^\text{th}$ partial sum of the odd series is $1+\frac13+\frac15+\cdots+\frac{1}{2n-1}$. The $n^\text{th}$ partial sum of the even series is $\frac12+\frac14 + \frac16 + \cdots +\frac1{2n}$. Each term of the even series is less than the corresponding term of the odd series, giving us our result.
\item		The $n^\text{th}$ partial sum of the odd series is $1+\frac13+\frac15+\cdots+\frac1{2n-1}$. The $n^\text{th}$ partial sum of 1 plus the even series is $1+\frac12+\frac14+\cdots + \frac{1}{2(n-1)}$. Each term of the even series is now greater than or equal to the corresponding term of the odd series, with equality only on the first term. This gives us the result.
\item		If the odd series converges, the work done in (a) shows the even series converges also. (The sequence of the $n^\text{th}$ partial sum of the even series is bounded and monotonically increasing.) Likewise, (b) shows that if the even series converges, the odd series will, too. Thus if either series converges, the other does. 

Similarly, (a) and (b) can be used to show that if either series diverges, the other does, too.
\item		If both the even and odd series converge, then their sum would be a convergent series. This would imply that the Harmonic Series, their sum, is convergent. It is not. Hence each series diverges.
\end{enumerate}
}
